# Linear ~ shape/size

This study begins by asking which linear measure/s---if any---can be said to covary with Perdiz arrow point shape and size? To assess covariance, Procrustes-aligned **shape** and centroid **size** are used in a pair of two-block partial least-squares analysis with each linear (caliper collected) metric. 

## Load packages + data

```{r loadlinearshapesize, echo=TRUE, warning=FALSE}
# download most recent software version
#devtools::install_github("geomorphR/geomorph", ref = "Stable", build_vignettes = TRUE)
#devtools::install_github("mlcollyer/RRPP")

# load analysis packages
library(here)
library(StereoMorph)
library(geomorph)
library(ggplot2)
library(dplyr)
library(ggpubr)
library(wesanderson)

# read shape data and define number of sLMs
shapes <- readShapes("shapes")
shapesGM <- readland.shapes(shapes, 
                            nCurvePts = c(10,3,5,5,3,10))

# read qualitative data
qdata <- read.csv("qdata.perdiz.csv",
                  header = TRUE,
                  row.names = 1)

# add derived vars to data
# maximum blade length (derived)
qdata$maxbl <- qdata$maxl - qdata$maxstl
# maximum shoulder width (derived)
qdata$maxshw <- qdata$maxw - qdata$maxstw
```

## Generalised Procrustes Analysis

Landmark data were aligned to a global coordinate system [@RN11622;@RN11623;@RN11563], achieved through generalized Procrustes superimposition [@RN478] performed in R 4.1.3 [@R] using the `geomorph` library v. 4.0.3 [@RN11530;@RN1774;@RN9565]. Procrustes superimposition translates, scales, and rotates the coordinate data to allow for comparisons among objects [@RN11564;@RN478]. The `geomorph` package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space subsequent to alignment in preparation for the use of multivariate methods that assume linear space [@RN1646;@RN11563]. 

```{r gpashapesize, echo=TRUE, out.width = "100%", dpi = 300, warning=FALSE}
# gpa
Y.gpa <- gpagen(shapesGM, print.progress = FALSE)

## plot
plot(Y.gpa)

# dataframe
gdf <- geomorph.data.frame(shape = Y.gpa$coords,
                           size = Y.gpa$Csize)
```

## 2BPLS Maximum length

### Shape

```{r 2bplsshl, echo=TRUE, out.width = "100%", dpi = 300, warning=FALSE}
# is Perdiz arrow point shape correlated with linear var?
shapeml <- two.b.pls(Y.gpa$coords,
                      qdata$maxl,
                      iter = 9999,
                      seed = NULL,
                      print.progress = FALSE)
summary(shapeml)

## plot
plot(shapeml)
```

### Size

```{r 2bplsszl, echo=TRUE, out.width = "100%", dpi = 300, warning=FALSE}
# is Perdiz arrow point size correlated with linear var?
sizeml <- two.b.pls(Y.gpa$Csize,
                      qdata$maxl,
                      iter = 9999,
                      seed = NULL,
                      print.progress = FALSE)
summary(sizeml)

## plot
plot(sizeml)
```

## 2BPLS Maximum blade length

### Shape

```{r 2bplsshbl, echo=TRUE, out.width = "100%", dpi = 300, warning=FALSE}
# is Perdiz arrow point shape correlated with linear var?
shapembl <- two.b.pls(Y.gpa$coords,
                      qdata$maxbl,
                      iter = 9999,
                      seed = NULL,
                      print.progress = FALSE)
summary(shapembl)

## plot
plot(shapembl)
```

### Size

```{r 2bplsszbl, echo=TRUE, out.width = "100%", dpi = 300, warning=FALSE}
# is Perdiz arrow point size correlated with linear var?
sizembl <- two.b.pls(Y.gpa$Csize,
                      qdata$maxbl,
                      iter = 9999,
                      seed = NULL,
                      print.progress = FALSE)
summary(sizembl)

## plot
plot(sizembl)
```

## 2BPLS Maximum shoulder width

### Shape

```{r 2bplsshshw, echo=TRUE, out.width = "100%", dpi = 300, warning=FALSE}
# is Perdiz arrow point shape correlated with linear var?
shapemshw <- two.b.pls(Y.gpa$coords,
                      qdata$maxshw,
                      iter = 9999,
                      seed = NULL,
                      print.progress = FALSE)
summary(shapemshw)

## plot
plot(shapemshw)
```

### Size

```{r 2bplsszshw, echo=TRUE, out.width = "100%", dpi = 300, warning=FALSE}
# is Perdiz arrow point size correlated with linear var?
sizemshw <- two.b.pls(Y.gpa$Csize,
                      qdata$maxshw,
                      iter = 9999,
                      seed = NULL,
                      print.progress = FALSE)
summary(sizemshw)

## plot
plot(sizemshw)
```

## 2BPLS Maximum width

### Shape

```{r 2bplsshw, echo=TRUE, out.width = "100%", dpi = 300, warning=FALSE}
# is Perdiz arrow point shape correlated with linear var?
shapemw <- two.b.pls(Y.gpa$coords,
                      qdata$maxw,
                      iter = 9999,
                      seed = NULL,
                      print.progress = FALSE)
summary(shapemw)

## plot
plot(shapemw)
```

### Size

```{r 2bplsszw, echo=TRUE, out.width = "100%", dpi = 300, warning=FALSE}
# is Perdiz arrow point size correlated with linear var?
sizemw <- two.b.pls(Y.gpa$Csize,
                      qdata$maxw,
                      iter = 9999,
                      seed = NULL,
                      print.progress = FALSE)
summary(sizemw)

## plot
plot(sizemw)
```

## 2BPLS Maximum thickness

### Shape

```{r 2bplsshth, echo=TRUE, out.width = "100%", dpi = 300, warning=FALSE}
# is Perdiz arrow point shape correlated with linear var?
shapemth <- two.b.pls(Y.gpa$coords,
                      qdata$maxth,
                      iter = 9999,
                      seed = NULL,
                      print.progress = FALSE)
summary(shapemth)

## plot
plot(shapemth)
```

### Size

```{r 2bplsszth, echo=TRUE, out.width = "100%", dpi = 300, warning=FALSE}
# is Perdiz arrow point size correlated with linear var?
sizemth <- two.b.pls(Y.gpa$Csize,
                      qdata$maxth,
                      iter = 9999,
                      seed = NULL,
                      print.progress = FALSE)
summary(sizemth)

## plot
plot(sizemth)
```

## 2BPLS Maximum stem length

### Shape

```{r 2bplsshstl, echo=TRUE, out.width = "100%", dpi = 300, warning=FALSE}
# is Perdiz arrow point shape correlated with linear var?
shapemstl <- two.b.pls(Y.gpa$coords,
                      qdata$maxstl,
                      iter = 9999,
                      seed = NULL,
                      print.progress = FALSE)
summary(shapemstl)

## plot
plot(shapemstl)
```

### Size

```{r 2bplsszstl, echo=TRUE, out.width = "100%", dpi = 300, warning=FALSE}
# is Perdiz arrow point size correlated with linear var?
sizemstl <- two.b.pls(Y.gpa$Csize,
                      qdata$maxstl,
                      iter = 9999,
                      seed = NULL,
                      print.progress = FALSE)
summary(sizemstl)

## plot
plot(sizemstl)
```

## 2BPLS Maximum stem width

### Shape

```{r 2bplsshstw, echo=TRUE, out.width = "100%", dpi = 300, warning=FALSE}
# is Perdiz arrow point shape correlated with linear var?
shapemstw <- two.b.pls(Y.gpa$coords,
                      qdata$maxstw,
                      iter = 9999,
                      seed = NULL,
                      print.progress = FALSE)
summary(shapemstw)

## plot
plot(shapemstw)
```

### Size

```{r 2bplsszstw, echo=TRUE, out.width = "100%", dpi = 300, warning=FALSE}
# is Perdiz arrow point size correlated with linear var?
sizemstw <- two.b.pls(Y.gpa$Csize,
                      qdata$maxstw,
                      iter = 9999,
                      seed = NULL,
                      print.progress = FALSE)
summary(sizemstw)

## plot
plot(sizemstw)
```
